---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , eval = F)
```

# Instructions
 - Clear the environment
 - Open a new R Script called `day5_exercise_script` where you will do the exercise and later save in the `day5` project directory.
 - Add the purpose of the file and the author [MANDATORY]
 - Here are the main activities for this excercise 
   1) Load the `rio`, `lubridate`,`epikit`, `janitor`, `infer` and `tidyverse` package
   2) Load the *WHR2018.csv* dataset using the `import` function.
   3) clean the column names to remove spaces
   4) Explore the distribution of all continuous variables using density, bar and boxplots

### 1 Importing the dataset into R and clean names

 - Download the *WHR2018.csv* from your emails and save it in the `Data` folder
 - Import the dataset into R using the `import` function
 - Clean the all column names to remove spaces betwen the names


### 2 Subsetting

 - Subset the imported dataset and only keep the following variables:
    - country
    - year
    - freedom_to_make_life_choices
    - confidence_in_national_government
    - positive_affect
    - negative_affect


### 3 Dropping records with missing data

 - In all the columns selected above, remove records that have missing data.


### 4 Scatter plot, correlation coefficient and line of best fit

 - Make a scatter plot of `negative_affect` vs. `positive_affect`
 - Calculate and interpret the correlation coefficient of `negative_affect` vs. `positive_affect`
 - Add a line of best fit.  *HINT:* add a layer of `geom_smooth()`



### 5 Linear regression 
 - Fit a linear regression model on `negative_affect` vs. `positive_affect`
 - Interpret the output


### 6 Chi square test
 - Do a `chi-square` test of facility utilization Pre-covid and during covid
 - Use the covid_tab table below
 - What do you think of the association?

```{r}
## CHi square covid
covid_tab <- matrix(
  c(3819, 7076, 215, 1261, 2390, 2345), 
    ncol=2,  byrow=F ,
  dimnames = list(c("Maternal service utilization", "Ooutpatient Utilization" ,"Inpatient utilization"),
                  c("Pre COVID", "During COVID")))

covid_tab

```


