---
title: "Pwani University R DAAD Training"
author: "Author: Moses Kioko"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    theme: united
    hightlight: tango
    fontsize: 24pt
    code_folding: show
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    fig_crop: false
    toc: no
    toc_depth: '6'
always_allow_html: yes 
editor_options: 
  chunk_output_type: inline
---

# Introduction to Numerical Analysis

In this section we be looking into **part** of numerical analysis of data which involves calculating;

+ Minimum Value 
+ Maximum Value 
+ Mean 
+ Median 
+ Standard Deviation 
+ Variance 
+ Moments 
  - First Moment 
  - Second Moment *(Mean)*
  - Third Moment *(Skewness)*
  - Fourth Moment *(Kurtosis)*

**Set Working Directory**

Setting up working environment entails clearing the working space and setting a computer directory from where you can launch scripting of your projects comfortably

```{r}
#clear working space 
rm(list =ls(all =T))

#set working directory 
setwd('D:/DAAD Training/Rmarkdown Script')
```


**Installing Libraries**

To access the functions hosted in package we need to install the package necessary for handling our data-set. The following are the packages names we are essential in this endeavor 


Package Name     |  Description 
|-------------------------|-----------------------------------------------------------------------------------------------
| Tidyverse Package       | Cleaning *(dplyr & tidyr)*, Manipulating*(dplyr & tidyr)* and plotting *(ggplot2)* data 
| Janitor Package         | Cleaning column names
| Lubridate Package       | Formatting and cleaning dates
| Reactable Package       | Printing out tables
| Reactablefmtr Package   | Formatting tables by adding conditional formatting elements 


**Please note:** *Our Main Package is Tidyverse the rest are just but addons*

```{r echo=TRUE, message=FALSE, warning=FALSE}
#installing libraries 
#install.packages('tidyverse',dependencies = T)
#install.packages('janitor',dependencies = T)
#install.packages('reactable', dependencies = T)
#install.packages('reactablefmtr',dependencies = T)

# install all the package at a go 
#install.packages(c('tidyverse','janitor','reactable','reactablefmtr'),dependencies = T)
```

**Libraries**

To make use of installed package we load them into R work-space using **Library()** command 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#Read data, manipulate and plot the data 
library(tidyverse)

#clean the column names 
library(janitor)

#calculating percentages
library(scales)

#table packages 
library(reactable)
library(reactablefmtr)
```

**Load Data-Set**

Load or read data are used interchangeably, we will read;

+ Child Weight Data-set
+ Child Death Data-set
```{r echo=TRUE, message=FALSE, warning=FALSE}
#read the data 
birth_weight <- readr::read_csv('birthweight2.csv')
child_death <- readr::read_csv('child.deaths.csv')
```

# Birth Weight Data-Set

## Numerical Analysis 
The first attempt on data-set should be to check the following 

+ Completeness - *missing values*
+ Duplicates - *Repeated record*
+ Variable Names - *do they conform to right format*
+ Data Types - *Numeric, characters/strings*

### Cleaning Variable/Column Names
 
There several ways on cleaning the column names by;

+ Replacing spaces with underscores 
+ Unify the variable name cases 
+ Replacing special characters with underscore

```{r}
birth_weight <- birth_weight %>% 
  janitor::clean_names()

child_death <- child_death %>% 
  janitor::clean_names()
```

### Birth Weight Duplicates 

Repeated records skew analysis , so to evade inaccurate analysis we check out the duplicated records.In R we can do it in two ways;

+ base package - *use duplicated in conjunction with which commands*
+ dplyr package - *use duplicate , duplicate_at ... etc*

```{r}
#flag the duplicates 
birth_index <- which(duplicated(birth_weight$id, fromLast = T))

#filter the data 
birth_weight_duplicate <- birth_weight[birth_index,]
birth_weight_nodup <- birth_weight[-birth_index,]
```

The ID for the duplicated record in birth weight data is **`r birth_weight_duplicate$id`** . We exclude one of this record before proceeding with our analysis.

### Base R Numerical Exploration 
We use base R commands to run descriptive analysis for numerical variables, this tests include calculating ; **Minimum**, **Maximum**,**Mean**, **Median**,**Variance**, **Standard Deviation** ..etc

```{r}
#calculate the maximum weight 
max_weight <- max(birth_weight_nodup$bweight, na.rm = T)

#calculate the minimum weight 
min_weight <- min(birth_weight_nodup$bweight, na.rm = T)

#calculate the mean weight 
mean_weight <- mean(birth_weight_nodup$bweight, na.rm = T)

#calculate the median weight 
median_weight <- median(birth_weight_nodup$bweight, na.rm = T)

#calculate the standard deviation weight 
sd_weight <- sd(birth_weight_nodup$bweight, na.rm = T)

#calculate the variance weight 
var_weight <- var(birth_weight_nodup$bweight, na.rm = T)

#calculate inter quatile ranges 
iqr_weight <- quantile(birth_weight_nodup$bweight, na.rm = T)

#calculate the range 
range_weight <- paste(min_weight,max_weight, sep = ' to ')
```

The following are the descriptive analysis for birth weight:

+ Minimum Birth Weight **`r min_weight`** grams 
+ Maximum Birth Weight **`r max_weight`** grams
+ Birth Weight Range **`r range_weight`** grams
+ Mean Birth Weight **`r round(mean_weight,2)`** grams
+ Median Birth Weight **`r round(median_weight,2)`** grams
+ Birth Weight Standard Deviation **`r round(sd_weight,2)`** grams
+ Birth Weight Inter Quantile Range  **`r iqr_weight`** grams

Most of the above single calculations can also be achieved using **summary()** command.

*Please note that I have only subset the data using square brackets to capture numeric values in data apart from sex variable*
```{r}
summary(birth_weight_nodup[,2:6])
```

### Tidyverse Numerical Exploration 

Tidyverse package comes in handy in handling data especially running numeric analysis. In this exercise we will dwell on **Dplyr Package** to tidy-up the data and run descriptive analysis of desired variables i.e **Birth Weight** ... etc

**Birth Weight Summaries for Child Gender **

Create descriptive analysis on birth weight distributed over gender 
```{r}
# Rename some variable names and lre-label 
birth_weight_data <- birth_weight_nodup %>% 
      dplyr::mutate(ethnic = case_when(ethnic==1 ~ 'group I',
                                     ethnic ==2 ~ 'group II',
                                     ethnic ==3 ~ 'group III',
                                     TRUE ~ 'group IV')) %>% 
  dplyr::rename(Gender =sex,
                `Age Group` = agegrp,
                `Weight Classification` =lbw,
                Ethnic = ethnic) 
#create summaries 
gender_weight <- birth_weight_data %>% 
  group_by(Gender) %>% 
  summarise(`Max Weight` = max(bweight,na.rm = T),
            `Min Weight` = min(bweight, na.rm = T),
            `Median Weight` = median(bweight, na.rm = T),
            `Mean Weight` = round(mean(bweight, na.rm = T),1),
            `Standard Deviation`= round(sd(bweight, na.rm = T),2)) 
```

**Simple Reactable Table Formart**

Print the output in a **simple table** with help of **reactable package**
```{r}
reactable(gender_weight)
```



**Advanced Reactable Table**

The customized  tables are easy to scan through since we highlight the maximum values and minimum values for each variable. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
reactable(gender_weight,
          striped = T,
          bordered = T,
          compact = T,
          columns = list(
            `Max Weight` = colDef(align = 'center',style  = highlight_min_max(gender_weight)),
            `Min Weight` = colDef(align ='center' ,style = highlight_min_max(gender_weight)),
            `Median Weight` = colDef(align ='center', style = highlight_min_max(gender_weight)),
            `Mean Weight` = colDef(align ='center', style = highlight_min_max(gender_weight))
          ))
```

### Age Group Weight Numerical Analysis 

This section highlights numerical analysis for weight grouped by age group

```{r echo=TRUE, message=FALSE, warning=FALSE}
#create summaries for agegroup 
agegroup_weight <- birth_weight_data %>% 
  group_by(`Age Group`) %>% 
  summarise(`Max Weight` = max(bweight,na.rm = T),
            `Min Weight` = min(bweight, na.rm = T),
            `Median Weight` = median(bweight, na.rm = T),
            `Mean Weight` = round(mean(bweight, na.rm = T),1),
            `Standard Deviation`= round(sd(bweight, na.rm = T),2)) 

#print summaries
reactable(agegroup_weight,
          striped = T,
          bordered = T,
          compact = T,
          columns = list(
            `Max Weight` = colDef(align = 'center',style  = highlight_min_max(agegroup_weight)),
            `Min Weight` = colDef(align ='center' ,style = highlight_min_max(agegroup_weight)),
            `Median Weight` = colDef(align ='center', style = highlight_min_max(agegroup_weight)),
            `Mean Weight` = colDef(align ='center', style = highlight_min_max(agegroup_weight))
          ))
```

### Ethic Group Gender and Weight Anaysis 

This section highlight weight numerical exploration grouped by gender and ethnic group.
```{r echo=TRUE, message=FALSE, warning=FALSE}
#create the summaries
gender_ethic_weight <- birth_weight_data%>% 
  group_by(Ethnic,Gender) %>% 
  summarise(Count = n(),
            `Max Weight` = max(bweight,na.rm = T),
            `Min Weight` = min(bweight, na.rm = T),
            `Median Weight` = median(bweight, na.rm = T),
            `Mean Weight` = round(mean(bweight, na.rm = T),1)) 

#print the table for gender ethnic and weight analysis
reactable(gender_ethic_weight,
          striped = T,
          bordered = T,
          compact = T,
          columns = list(
            `Max Weight` = colDef(align = 'center',style  = highlight_min_max(gender_ethic_weight)),
            `Min Weight` = colDef(align ='center' ,style = highlight_min_max(gender_ethic_weight)),
            `Median Weight` = colDef(align ='center', style = highlight_min_max(gender_ethic_weight)),
            `Mean Weight` = colDef(align ='center', style = highlight_min_max(gender_ethic_weight)),
            Count = colDef(align ='center', style = highlight_min_max(gender_ethic_weight))
          ))
```

# Child Death Data-Set

+ Labeling gender(sex)
+ Formatting date 
  - Calculate the date differences
  
```{r}
child_death_clean <- child_death %>% 
  dplyr::mutate(sex = case_when(sex =='m' ~ 'Male', #change the labels 
                                TRUE ~ 'Female')) %>% 
  dplyr::mutate(date_birth = lubridate::mdy(date_birth)) %>% # change character to date formart 
  dplyr::mutate(date_exit = lubridate::mdy(date_exit)) %>% 
  dplyr::mutate(days= difftime(date_exit,date_birth, units ='days')) %>%  #calculate the live span
  dplyr::mutate(days= as.numeric(days)) %>%  # convert to numeric
  dplyr::mutate(mom_educ = str_to_title(mom_educ)) %>%  #change the case for mother education 
  dplyr::rename(`Mom Education` = mom_educ, #rename variables names
                `Birth Location` = locn_birth,
                Gender = sex) %>% 
  dplyr::mutate(status = case_when(status==1 ~ 'Status I',
                                   TRUE ~ 'Status II')) 
```

## Child Death Descriptive Analysis
This section we will be calculating the number of children and group them with the location of birth and level of mothers education.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# renaming multiple columns
old_names <- c('status','Home','Hospital')
new_names <- c('Status','No: Home','No: Hospital')

#summaries 
place_birth <- child_death_clean %>% 
  group_by(`Mom Education`,`Birth Location`,status) %>% 
  summarise(Count = n()) %>% 
  #calculate percentages 
  pivot_wider(names_from = `Birth Location`,values_from = Count) %>%
  dplyr::mutate(`Hospital Delivery` = Hospital/(sum(Home)+sum(Hospital))) %>% 
  #formart the percentages 
  dplyr::mutate(`Hospital Delivery` = scales::percent(`Hospital Delivery`,accuracy = 0.1)) %>% 
  #rename variables
  dplyr::rename_at(vars(old_names),~ new_names)



  

#print the table
reactable(place_birth,
          striped = T, 
          bordered =  T,
          columns = list(
            `No: Home` = colDef(align = 'center',style = highlight_min_max(place_birth)),
            `No: Hospital` = colDef(align = 'center', style = highlight_min_max(place_birth))
          ))
```

## Dates Analysis 

This section will be looking at calculating the live span of the children based on their gender and status.

```{r echo=TRUE, message=FALSE, warning=FALSE}
#create date summaries 
live_span <- child_death_clean %>% 
  dplyr::select_all(str_to_title) %>% 
  dplyr::group_by(`Birth Location`,Status) %>% 
  dplyr::summarise(`Min Days` = min(Days,na.rm = T),
                   `Max Days` = max(Days,na.rm =T),
                   `Average Days` = median(Days,na.rm = T)) 

#print the date summaries in table
reactable::reactable(live_span,
                     bordered = T,
                     striped = T,
                     compact = T,
                     columns = list(
                      `Min Days` = colDef(align = 'center',style = highlight_min(live_span)),
                      `Max Days`  = colDef(align ='center',style = highlight_min(live_span)),
                      `Average Days` = colDef(align = 'center',style = highlight_min(live_span))
                     ))
```

