---
title: "Exploratory analysis of categorical data "
author: "Alice Kamau"
date: "7/5/2021"
output: 
  slidy_presentation: default
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE,
      tidy.opts=list(blank=FALSE, width.cutoff=60,size = 'tiny'),
      fig.width=10, 
      fig.height=8)
library(tidyverse)
library(knitr)
```

## Learning objectives

- Explain key procedures for the analysis of categorical data

- Use R to perform tests on proportions for one, two or k categorical variables

- Interpret the results of tests on proportions for one, two or k categorical variables


## Understanding categorical variables

- When we calculate summaries of categorical variables we are aiming to describe the sample distribution of the variable, just as with numeric variables.

- The general question we need to address is, ‘what are the relative frequencies of different categories?’

- Since a categorical variable takes a finite number of possible values, the simplest thing to do is tabulate the number of occurances of each type.


- Load required package

```{r , eval=FALSE}
library(tidyverse)
```

- Set the directory

```{r, eval=FALSE}
setwd("/Users/akamau/Documents/I-StaR/Course 1/Day5/Presentation/")
```

- Load the data

```{r}
bw_df <- read.csv("Data/birthweight2.csv")

```

## table() & prop.table()

- table() & prop.table() is a quick way to pull together row/column frequencies and proportions for categorical variables

```{r}

table(bw_df$lbw2)

prop.table(table(bw_df$lbw2))

```

##

```{r}

ggplot(bw_df, aes(agegrp)) + 
    geom_bar(aes(y = (..count..)/sum(..count..))) + 
    scale_y_continuous(labels=scales::percent) +
    ylab("Proportion (%)")

```

## R functions: binom.test() & prop.test()

- The R functions binom.test() and prop.test() can be used to perform one-proportion test:

- binom.test(): compute exact binomial test. Recommended when sample size is small

- prop.test(): can be used when sample size is large ( N > 30). It uses a normal approximation to binomial

- The syntax of the two functions are exactly the same. The simplified format is as follow:

```{r,  eval=FALSE , class.source='highlight_code'}
# syntax
binom.test(x, n, p = 0.5, alternative = "two.sided")
prop.test(x, n, p = NULL, alternative = "two.sided",
          correct = TRUE)

x: the number of of successes
n: the total number of trials
p: the probability to test against.
correct: a logical indicating whether Yates’ continuity correction should be applied where possible.

```

## One sample proportion test

- One sample proportion test is used to compare an observed proportion to a theoretical one, when there are only two categories.

```{r}

# Assumming a normal approximation
# H0: Proportion of normal birthweight = 90%

prop.test(sum(bw_df$lbw2==0),length(bw_df$lbw2==0),p=.9,correct=T)

# Using exact binomial test

binom.test(sum(bw_df$lbw2==0),length(bw_df$lbw2==0),p=.9)
```

##

- The function returns:

    - the value of Pearson’s chi-squared test statistic.
    - a p-value
    - a 95% confidence intervals
    - an estimated probability of success (the proportion of children with normal weight)
    
## Recap of basic tools for analysing binary data

- Descriptive: Bar charts and tabulations

- Analytic: Use of prop.test() assuming normal approximation or using binom.test() based on the exact distribution


## Exercise

- Use birthweight2

- Check the variables, and explore the data.

- Generate a barplot of lbw stratified by sex

- Get the proportion of low birth weight babies and 95% CI.

- Get the proportion of lbw babies (and 95% CI) by sex.

- Test this hypothesis p=0.90 (90% normal BW) for female babies and male babies separately


## Solution

- Check the variables, and explore the data.

```{r}
names(bw_df)
str(bw_df)
```

##

```{r}

head(bw_df)

```

##

- Generate a barplot of lbw stratified by sex (default is stacked)

```{r}

 ggplot(bw_df, aes(x = lbw)) +
  geom_bar(aes(fill = sex))

```

##

- Generate a barplot of lbw stratified by sex (side by side)

```{r}
brks <- c(0, 0.25, 0.5, 0.75, 1)

bw_df %>% 
  group_by(sex, lbw) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) %>% 
ggplot(aes(x=lbw,y = perc*100,fill=sex)) + 
    geom_bar(stat="identity", width = 0.7,position="dodge") +
  labs(y="Proportion (%)", x="Birth weight") + theme_bw()

```

##

- Get the proportion of low birth weight babies and 95% CI.

```{r}
prop.test(sum(bw_df$lbw2==1),length(bw_df$lbw2==1))

```

## 

- Get the proportion of lbw babies (and 95% CI) by sex.

```{r}

table(bw_df$sex)

bw_female<-bw_df %>% 
  filter(sex=="Female") 

prop.table(table(bw_female$lbw2))

bw_male<-bw_df %>% 
  filter(sex=="Male") 

prop.table(table(bw_male$lbw2))

```

## Females

- Test this hypothesis p=0.90 (90% normal BW) for female babies and male babies separately

```{r}

prop.test(sum(bw_female$lbw2==0),length(bw_female$lbw2==0),p=.9,correct=T)
```

## Males

```{r}

prop.test(sum(bw_male$lbw2==0),length(bw_male$lbw2==0),p=.9,correct=T)
```

## Associations between categorical variables

- The general question we need to address is, “do different combinations of categories seem to be under or over represented?” 

- We need to understand which combinations are common and which are rare. 

- The simplest thing we can do is ‘cross-tabulate’ the number of occurrences of each combination. 

- The resulting table is called a contingency table. 

- The counts in the table are sometimes referred to as frequencies.


##

```{r}
tab1<- table(bw_df$sex,bw_df$lbw2)

prop.table(tab1, 1) ## row proportions
prop.table(tab1, 2) ## column proportions

##
prop.table(table(bw_df$sex,bw_df$lbw2),1)
```

## Using dplyr & tidyr: Crosstabs

- A good reasons for not just using the base table() command is when you dealing with missing data

```{r}

## frequency

bw_df%>%
  group_by(sex, lbw2)%>%
  summarise(n=n())%>%
  spread(lbw2, n)%>%
  kable()

```

##

```{r}

## proportion

bw_df%>%
  group_by(sex, lbw2)%>%
  summarize(n=n())%>%
  mutate(prop=n/sum(n))%>%
  subset(select=c("sex","lbw2","prop"))%>%   #drop the frequency value
  spread(lbw2, prop) %>%
  kable()

```

## Including some missing data

```{r}
bw_df %>%
  mutate(
    ethnic2 = ifelse(ethnic==2, NA, ethnic),
    agegrp2=ifelse(agegrp=="40+yrs",NA,agegrp)
  ) %>%
  group_by(ethnic2, agegrp2) %>%
  tally() %>%
  data.frame() %>% ### <-- go from tbl_df to data.frame
  spread(agegrp2, n)

```

## Using base R table() command

- base R ignores this missing value in the output table

```{r}
bw_df$ethnic2<- ifelse(bw_df$ethnic==2, NA, bw_df$ethnic)
bw_df$agegrp2<- ifelse(bw_df$agegrp=="40+yrs", NA, bw_df$agegrp)

table(bw_df$ethnic2,bw_df$agegrp2)
```

## Comparing proportions

- To test the hypothesis that the proportions are different, there are several ways to do this: 
    
    - Using chi-squared test 
    - Using fishers exact test
    - Two sample proportion test


## Chi-squared test - Comparing proportions 

- Comparing two (or more) proportions - the Chi-squared test uses Expected numbers.

- Chi-squared test is valid for any contingency table

- Assumptions: sufficient numbers in each cell of the table

    - State the null hypothesis: No association between the two variables.
    - Calculate the Chi-squared statistic from the Observed and Expected numbers
    - Obtain the p-value for the data, under H0

##

```{r}

chisq.test(bw_df$sex,bw_df$lbw2)

# If Chi-squared test not valid then get R to test the null hypothesis H0 using the Fishers exact test.

fisher.test(bw_df$sex,bw_df$lbw2)

```

## Two sample proportion test

```{r}

prop.test(table(bw_df$sex,bw_df$lbw2))

```

## Exercise

- Use birthweight2, with outcome low birth weight (lbw)

- Ensure you have the variable that shows 1= LBW, 0=Normal

- Compare the proportion with low birth weight by the ethnic groups. 

- Tabulate and test if lbw differs by ethnic. 

- Tabulate the low birth weight by hypertension status of mothers (variable is called ht) 

- Look at the association between lbw and hypertension (ht), using the chi-squared test


## Solutions

- Ensure you have the variable that shows 1= LBW, 0=Normal

```{r} 

table(bw_df$lbw2)

```

- Compare the proportion with low birth weight by the ethnic groups.

```{r}

prop.table(table(bw_df$ethnic,bw_df$lbw2),1)

```

## 

- Test if lbw differs by ethnic. 

```{r} 

## Chi-squared test can be used for larger tables, with more categories (e.g. ethnic).

chisq.test(bw_df$ethnic,bw_df$lbw2)

prop.test(table(bw_df$ethnic,bw_df$lbw2))

```

## 

- Tabulate the low birth weight by hypertension status of mothers (variable is called ht) 

```{r}

## frequency

bw_df%>%
  group_by(ht, lbw2)%>%
  summarise(n=n())%>%
  spread(lbw2, n)%>%
  kable()

```

##

```{r}

## proportion

bw_df%>%
  group_by(ht, lbw2)%>%
  summarize(n=n())%>%
  mutate(prop=n/sum(n))%>%
  subset(select=c("ht","lbw2","prop"))%>%   #drop the frequency value
  spread(lbw2, prop)%>%
  kable()

```

##

- Look at the association between lbw and hypertension (ht), using the chi-squared test

```{r}

chisq.test(bw_df$ht,bw_df$lbw2)

prop.test(table(bw_df$ht,bw_df$lbw2))

```

