---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , eval = F)
```

# Instructions
 - Clear the environment
 - Open a new R Script called `day4_exercise_script` where you will do the exercise and later save in the `day4` project directory.
 - Add the purpose of the file and the author [MANDATORY]
 - Here are the main activities for this excercise 
   1) Load the `rio`, `lubridate`,`epikit`, `janitor`, `infer` and `tidyverse` package
   2) Load the blood *class presusure data* from Mohammed's study. Use the `import` function.
   3) Clean a string variable
   4) Explore the distribution of all continuous variables using density, bar and boxplots
   5) Plot frequency bar plots
   6) Compare means and proportions
   


### 1 Importing the dataset into R

 - Save the *BP_measure.xlsx* in the `Data` folder
 - Import the dataset into R using the `import` function



### 2 Data cleaning

 - Clean the `syst_diast` variable to separate the diastolic and systolic measurements.
 - Create a new variable (name it `hr_cat`) that classifies `hr` as *high* or *low* using a cutt-off of 72.


### Plotting graphs

 - Make a *boxplot* that compares `hr` by `age_cat`
 - Make a *boxplot* that compares `hr` by `sex`. Make this one horizontal.
 - Make a *bar graph* that shows the fequencies of male and female respondents
 


### Data summaries

 - Make data summaries that shows the mean `hr` by sex. Hint: use `group_by()` and `summarise()`
 - Make data summaries that shows the mean `hr` by age_cat. 


### Comparison of means and proportions

 - Compare mean `hr` between males and females and interpret.
 
 *HINT:* 
 **Helper code**  
```{r 1.1response, eval=FALSE}
# example of comparing means

t.test(data=dataframe ,
       hr_cat_variable ~ hr_cat_variable)
```
 
 
 - Compare proportions of or respondents with high `hr` in males and females. Use the variable `hr_cat`. Interpret.
 
 *HINT:* 
 **Helper code**  
```{r 1.2response, eval=FALSE}
# example of comparing proportions

infer::prop_test(dataframe ,
                 hr_cat_variable ~ sex)
```



### Reshaping

 - Use `select()` to select *age_cat*, *sex* and *hr* and put then in an object named *hr_sex*
 - Use `pivot_wider()` to reshape *hr_sex* and put the resulting dataset in an object named *hr_sex_wider*, where the new columns will be *male* and *female*
 
 *HINT:* 
 **Helper code**  
```{r 1.3response, eval=FALSE}
# example of reshaping from long to wide
hr_sex_wider <- hr_sex %>% 
  pivot_wider(names_from = ,
              values_from = )
```

